Appendix 3: Estimating species interactions

### 2015-10-25 Harris_2015/manuscript-materials/draft.tex
various

\textbf{Title:} Estimating species interactions from observational data
with Markov networks

@@

\textbf{Abstract:} Estimating species interactions from observational
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that is capable of
outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can
predict the direct and indirect consequences of any possible species
interaction matrix. Interactions in these models can be estimated from
observational data via maximum likelihood. Using simulated landscapes
with known pairwise interaction strengths, I evaluated Markov networks
and six existing approaches. The Markov networks consistently
outperformed other methods, correctly isolating direct interactions
between species pairs even when indirect interactions largely
overpowered them. Two computationally efficient approximations, based on
linear and generalized linear models, also performed well. Indirect
effects reliably caused a common null modeling approach to produce
incorrect inferences, however.

@@

To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et
al. 2013), existing methods for detecting competition's effects on
community structure are unreliable (Gotelli and Ulrich 2009). In
particular, species' effects on one another can become lost in the
complex web of direct and indirect interactions in real assemblages. For
example, the competitive interaction between the two shrub species in
Figure 1A can become obscured by their shared tendency to occur in
unshaded areas (Figure 1B). While ecologists have long known that
indirect effects can overwhelm direct ones at the landscape level
(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
inferenes from observational data do not control for these effects (e.g.
Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013,
Pollock et al. 2014). To the extent that indirect interactions like
those in Figure 1 are generally important (Dodson 1970), existing
methods will thus not generally provide much evidence regarding species'
direct effects on one another. The goal of this paper is to resolve this
long-standing problem.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most of our methods assume), it nevertheless
does leave a signal in the data (Figure 1C). Specifically, after
partitioning the data set into shaded sites and unshaded sites, there
will be co-occurrence deficits in each subset that might not be apparent
at the landscape level. More generally, we can obtain much better
estimates of the association between two species from their conditional
relationships (i.e.~by controlling for other species in the network)
than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumption that all the pairwise interactions are zero and thus don't
need to be controlled for. Nevertheless, null models have dominated this
field for more than three decades (Strong et al. 1984, Gotelli and
Ulrich 2009).

Following recent work by Azaele et al. (2010) and Fort (2013), this
paper shows that Markov networks (undirected graphical models also known
as Markov random fields; Murphy 2012) can provide a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observed presence-absence
matrices. Markov networks have been used in many scientific fields in
similar contexts for decades, from physics (where nearby particles
interact magnetically; Cipra 1987) to spatial statistics (where adjacent
grid cells have correlated values; Harris 1974, Gelfand et al. 2005).
While community ecologists explored some related approaches in the
1980's (Whittam and Siegel-Causey 1981), they used severe approximations
that led to unintelligible results (e.g. ``probabilities'' greater than
one; Gilpin and Diamond 1982).

Below, I introduce Markov networks and show how they can be used to
simulate landscape-level data or to make exact predictions about the
direct and indirect consequences of possible interaction matrices. Then,
using simulated data sets where the ``true'' interactions are known, I
compare this approach with several existing methods. Finally, I discuss
opportunities for extending the approach presented here to other
problems in community ecology, e.g.~quantifying the overall effect of
species interactions on occurrence rates (Roughgarden 1983) and
disentangling the effects of biotic versus abiotic interactions on
species composition (Kissling et al. 2012, Pollock et al. 2014).

@@

Markov networks provide a framework for translating back and forth
between the conditional relationships among species (Figure 1C) and the
kinds of species assemblages that these relationships produce. Here, I
show how a set of conditional relationships can be used to determine how
groups of species can co-occur. Methods for estimating conditional
relationships from data are discussed in the next section.

@@

\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\)

@@

Here, \(\alpha_{i}\) is an intercept term determining the amount that
the presence of species \(i\) contributes to the log-probability of
\(\vec{y}\); it directly controls the prevalence of species \(i\).
Similarly, \(\beta_{ij}\) is the amount that the co-occurrence of
species \(i\) and species \(j\) contributes to the log-probability; it
controls the conditional relationship between two species, i.e.~the
probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). For example,
\(\beta_{ij}\) might have a value of \(+2\) for two mutualists,
indicating that the odds of observing one species are \(e^2\) times
higher in sites where its partner is present than in comparable sites
where its partner is absent. Because the relative probability of a
presence-absence vector increases when positively-associated species
co-occur and decreases when negatively-associated species co-occur, the
model tends---all else equal---to produce assemblages that have many
pairs of positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from
this expectation. Determining the marginal relationships between species
from their conditional interactions entails summing over the different
possible assemblages (Figure 2B). This becomes intractable when the
number of possible assemblages is large, though several methods beyond
the scope of this paper can be employed to keep the calculations
feasible (Salakhutdinov 2008, Lee and Hastie 2012). Alternatively, as
noted below, some common linear and generalized linear methods can also
be used as computationally efficient approximations to the full network
(Lee and Hastie 2012, Loh and Wainwright 2013).

@@

In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a matrix of ones and zeros indicating which species are present at which
sites). When the number of species is reasonably small, one can compute
exact maximum likelihood estimates for all of the \(\alpha\) and
\(\beta\) coefficients given a presence-absence matrix by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will always converge on the global maximum.
This maximum represents the unique combination of \(\alpha\) and
\(\beta\) coefficients that would be expected to produce exactly the
observed co-occurrence frequencies on average (i.e.~maximizing the
likelihood matches the sufficient statistics of the model distribution
to the sufficient statistics of the data; Murphy 2012). I used the
rosalia package (Harris 2015a) for the R programming language (R Core
Team 2015) to optimize the Markov network parameters. The package was
named after Santa Rosalia, the patron saint of biodiversity, whose
supposedly miraculous healing powers played an important rhetorical role
in the null model debates of the 1970's and 1980's (Lewin 1983).

@@

In order to compare different methods, I simulated two sets of
landscapes using known parameters. The first set included the three
competing species shown in Figure 1. For each of 1000 replicates, I
generated a landscape with 100 sites by sampling from a probability
distribution defined by the figure's interaction coefficients (Appendix
1). Each of the methods described below was then evaluated on its
ability to correctly infer that the two shrub species competed with one
another, despite their frequent co-occurrence.

I also simulated a second set of landscapes using a stochastic community
model based on generalized Lotka-Volterra dynamics, as described in
Appendix 2. In these simulations, each species pair was randomly
assigned to either compete for a portion of the available carrying
capacity (negative interaction) or to act as mutualists (positive
interaction). Here, mutualisms operate by mitigating the effects of
intraspecific competition on each partner's death rate. For these
analyses, I simulated landscapes with up to 20 species and 25, 200, or
1600 sites (50 replicates per landscape size; see Appendix 2).

@@

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov newtork
models, as described above. For the analyses with 20 species, I added a
very weak logistic prior distribution on the \(\alpha\) and \(\beta\)
terms with scale 2 to ensure that the model estimates were always
finite. The bias introduced by this prior should be small: the 95\%
credible interval on \(\beta\) only requires that one species' effect on
the odds of observing a different species to be less than a factor of
1500 (which is not much of a constraint). The logistic distribution was
chosen because it is convex and has a similar shape to the Laplace
distribution used in LASSO regularization (especially in the tails), but
unlike the Laplace distribution it is differentiable everywhere and does
not force any estimates to be exactly zero. To confirm that this
procedure produced stable estimates, I compared its estimates on 50
bootstrap replicates (Appendix 4).

I also evaluated six alternative methods: five from the existing
literature, plus a novel combination of two of these methods. The first
alternative interaction metric was the sample correlation between
species' presence-absence vectors, which summarizes their marginal
association. Next, I used partial correlations, which summarize species'
conditional relationships (Albrecht and Gotelli 2001, Faisal et al.
2010). In the context of non-Gaussian data, the partial correlation can
be thought of as a computationally efficient approximation to the full
Markov network model (Loh and Wainwright 2013). This sort of model is
very common for estimating relationships among genes and gene products
(Friedman et al. 2008). Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor
package's \texttt{pcor.shrink} function with the default settings
(Schäfer et al. 2014).

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network (Lee and Hastie 2012). Following Faisal et al. (2010), I fit
regularized logistic regression models (Gelman et al. 2008) for each
species, using the other species on the landscape as predictors. To
avoid the identifiability problems associated with directed cyclic
graphs (Schmidt and Murphy 2012), I then symmetrized the relationships
within species pairs via averaging.

@@

I then evaluated the relationship between each method's estimates and
the ``true'' interaction strengths among all of the species pairs from
the larger simulated landscapes. This determined which of the methods
provide a consistent way to know how strong species interactions
are---regardless of which species were present in a particular data set
or how many observations were taken. Because the different methods
mostly describe species interactions on different scales
(e.g.~correlations versus \(Z\) scores versus regression coefficients),
I used linear regression through the origin to rescale the different
estimates produced by each method so that they had a consistent
interpretation. After rescaling each method's estimates, I calculated
squared errors between the scaled interaction estimates and ``true''
interaction values across all the simulated data sets. These squared
errors determined the proportion of variance explained for different
combinations of model type and landscape size (compared with a null
model that assumed all interaction strengths to be zero).

@@

Despite some variability across contexts (Figure 3A), the four methods
that controlled for indirect effects clearly performed the best: the
Markov network explained the largest portion of the variance in the
``true'' interaction coefficients (35\% overall), followed by the
generalized linear models (30\%), partial correlations from the raw
presence-absence data (28\%), and partial correlations from BayesComm,
the joint species distribution model (26\%). The benefit of choosing the
full Markov network over the other three methods was largest on the
smaller landscapes, which are also the ones that are most representative
of typical analyses in this field (Gotelli and Ulrich 2009).

The three methods that did not attempt to control for indirect
interactions all explained less than 20\% of the variance. Of these, the
sample correlation matrix based on the raw data performed the best
(19\%), followed by the null model (15\%) and BayesComm's correlation
matrix (11\%). Although these last three methods had different \(R^2\)
values, there was a close mapping among their estimates (especially
after controlling for the size of the simulated landscapes; Figure 3B).
This suggests that the effect sizes from the null model (and, to a
lesser extent, the correlation matrices from joint species distribution
models) only contain noisy versions of the same information that could
be obtained more easily and interpretably by calculating correlation
coefficients between species' presence-absence vectors.

Bootstrap resampling indicated that the above ranking of the different
methods was robust (Appendix 3). In particular, the 95\% confidence
interval of the bootstrap distribution indicated that the Markov
network's overall \(R^2\) value was between 14 and 18 percent higher
than the second-most effective method (generalized linear models) and
between 2.12 and 2.38 times higher than could be achieved by the null
model (Pairs). Bootstrap resampling of a 200-site landscape also
confirmed that the rosalia package's estimates of species' conditional
relationships were robust to sampling variation for reasonably-sized
landscapes (Appendix 4).

@@

The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where a common null modeling technique reliably fails.
Specifically, Markov networks were successful even when direct
interactions were largely overwhelmed by indirect effects (Figure 1).
For cases where fitting a Markov network is computationally infeasible,
these results also indicate that partial covariances and generalized
linear models (the two methods that estimated conditional relationships
rather than marginal ones) can both provide useful approximations. The
partial correlations' success on simulated data may not carry over to
real data sets, however; Loh and Wainwright (2013) show that the linear
approximations can be less reliable in cases where the true interaction
matrix contains more structure (e.g.~guilds or trophic levels).
Similarly, the approximation involved in using separate generalized
linear models for each species can occasionally lead to catastrophic
overfitting with small-to-moderate sample sizes (Lee and Hastie 2012).
For these reasons, it will usually be best to fit a Markov network
rather than one of the alternative methods when one's computational
resources allow it.

@@

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century (Cipra 1987), and the
models' properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species
interactions would thus allow ecologists to tap into an enormous set of
existing discoveries and techniques for dealing with indirect effects,
stability, and alternative stable states. Numerous other extensions are
possible: for example, the states of the interaction network can be
modeled as a function of the local abiotic environment (Lee and Hastie
2012), which would provide a rigorous and straightforward approach to
the difficult and important task of incorporating whole networks of
biotic interactions into species distribution models (Kissling et al.
2012, Pollock et al. 2014), leading to a better understanding of the
interplay between biotic and abiotic effects on community structure.
There are even methods (Whittam and Siegel-Causey 1981, Tjelmeland and
Besag 1998) that would allow one species to affect the sign or strength
of the relationship between two other species, tipping the balance
between facilitation and exploitation (Bruno et al. 2003).

@@

@@

Fort, H. 2013. Statistical Mechanics Ideas and Techniques Applied to
Selected Problems in Ecology. Entropy 15:5237--5276.

@@

Gotelli, N. J., and G. L. Entsminger. 2003. Swap algorithms in null
model analysis. Ecology:532--535.

@@

Kissling, W. D., C. F. Dormann, J. Groeneveld, T. Hickler, I. Kühn, G.
J. McInerny, J. M. Montoya, C. Römermann, K. Schiffers, F. M. Schurr, A.
Singer, J.-C. Svenning, N. E. Zimmermann, and R. B. O'Hara. 2012.
Towards novel approaches to modelling biotic interactions in
multispecies assemblages at large spatial extents. Journal of
Biogeography 39:2163--2178.

@@

Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.

\textbf{Figure captions}

\setlength{\parskip}{3pt} \setlength{\parindent}{0em}
\setlength{\leftskip}{0em}

\textbf{Figure 1.} \textbf{A.} A small network of three competing
species. The tree (top) tends not to co-occur with either of the two
shrub species, as indicated by the strongly negative coefficient linking
them. The two shrub species also compete with one another, but more
weakly (circled coefficient). \textbf{B.} In spite of the competitive
interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors
positively correlated (circled). \textbf{C.} Controlling for the tree
species' presence with a conditional method such as a partial covariance
or a Markov network leads to correct identification of the negative
shrub-shrub interaction (circled).

\textbf{Figure 2.} \textbf{A.} A small Markov network with two species.
The abiotic environment favors the occurrence of both species
(\(\alpha >0\)), particularly species 2 (\(\alpha_2 > \alpha_1\)). The
negative \(\beta\) coefficient linking these two species implies that
they co-occur less than expected under independence. \textbf{B.}
Relative probabilities of all four possible presence-absence
combinations for Species 1 and Species 2. The exponent includes
\(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but not when
it is absent (\(y_1 = 0\)). Similarly, the exponent includes
\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
constant \(Z\), ensures that the four relative probabilities sum to 1.
In this case, \(Z\) is about 18.5. \textbf{C.} We can find the expected
frequencies of all possible co-occurrence patterns between the two
species of interest. \textbf{D.} If \(\beta_{12}\) equaled zero (e.g.~if
the species no longer competed for the same resources), then the
reduction in competition would allow each species to increase its
occurrence rate and the co-occurrence deficit would be eliminated.

\textbf{Figure 3.} \textbf{A.} Proportion of variance in interaction
coefficients explained by each method versus number of sampled
locations. \textbf{B.} The \(Z\)-scores produced by the null model
(``Pairs'') for each pair of species can be predicted using the
correlation between the presence-absence vectors of those same species
and from the number of sites on the landscape.

\end{flushleft}
\end{spacing}

>>>>

\textbf{Title:} Estimating species interactions from co-occurrence data
with Markov networks

@@

\textbf{Abstract:} Estimating species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that is capable of
outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can
predict the direct and indirect consequences of any possible species
interaction matrix. Interactions in these models can also be estimated
from observed co-occurrence rates via maximum likelihood. Using
simulated landscapes with known pairwise interaction strengths, I
evaluated Markov networks and six existing approaches. The Markov
networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect
interactions largely overpowered them. Two computationally efficient
approximations, based on controlling for indirect effects with linear or
generalized linear models, also performed well. Indirect effects
reliably caused a common null modeling approach to produce incorrect
inferences, however.

@@

To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Connor et al. 2013), existing methods
for detecting competition's effects on community structure are
unreliable (Gotelli and Ulrich 2009). In particular, species' effects on
one another can become lost in the complex web of indirect effects. For
example, the competitive interaction between the two shrub species in
Figure 1A can become obscured by their shared tendency to occur in
unshaded areas (Figure 1B). While ecologists have long known that
indirect effects can overwhelm direct ones at the landscape level
(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
inferences from observational data do not control for these effects
(e.g. Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech
2013, Pollock et al. 2014). To the extent that indirect interactions
like those in Figure 1 are generally important (Dodson 1970), existing
methods will thus not generally provide much evidence regarding species'
direct effects on one another. The goal of this paper is to resolve this
long-standing problem.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most methods assume), it nevertheless does
leave a signal in the data (Figure 1C). For example, after partitioning
the data set into shaded sites and unshaded sites, there will be
co-occurrence deficits in each subset that might not be apparent at the
whole-landscape level. More generally, controlling for other species in
the network will often be important for obtaining reliable estimates of
direct (conditional, or all-else-equal) effects. This kind of precision
is difficult to obtain from null models, which only simulate the most
extreme possible scenario: that \emph{all} direct and indirect
interactions are exactly zero. Nevertheless, null models have dominated
this field for more than three decades (Strong et al. 1984, Gotelli and
Ulrich 2009).

Following Azaele et al. (2010), this paper shows that Markov networks
(undirected graphical models also known as Markov random fields; Murphy
2012) can provide a framework for understanding the landscape-level
consequences of pairwise species interactions, and for detecting them
from observed presence-absence matrices. Markov networks have been used
in many scientific fields in similar contexts for decades, from physics
(where nearby particles interact magnetically; Cipra 1987) to spatial
statistics (where adjacent grid cells have correlated values; Harris
1974, Gelfand et al. 2005). While community ecologists explored some
related approaches in the 1980's (Whittam and Siegel-Causey 1981), they
used severe approximations that led to unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982).

Below, I introduce Markov networks and show how they can be used to
simulate landscape-level data from an interaction matrix or to make
exact predictions about the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the ``true''
interactions are known, I compare this approach with several existing
methods. Finally, I discuss opportunities for extending the approach
presented here to other problems in community ecology, e.g.~quantifying
the overall effect of species interactions on occurrence rates
(Roughgarden 1983) and disentangling the effects of biotic versus
abiotic interactions on species composition (Pollock et al. 2014).

@@

Markov networks provide a framework for translating back and forth
between the conditional (all-else-equal) relationships among species
(Figure 1C) and the kinds of species assemblages that these
relationships produce. Here, I show how a set of conditional
relationships can be used to determine how groups of species can
co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),\)

@@

\noindent where the second sum is over all \(\binom{n}{2}\) species
pairs. In this model, \(\alpha_{i}\) is an intercept term determining
the amount that the presence of species \(i\) contributes to the
log-probability of \(\vec{y}\); it directly controls the prevalence of
species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it controls the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, \(\beta_{ij}\) might have a value of \(+2\) for two
mutualists, indicating that the odds of observing one species are
\(e^2\) times higher in sites where its partner is present than in
comparable sites where its partner is absent. Because the relative
probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else
equal---to produce assemblages that have many pairs of
positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' overall co-occurrence patterns can be more
complicated. Determining the marginal relationships between species from
their conditional interactions entails summing over the different
possible assemblages (Figure 2B). This becomes intractable when the
number of possible assemblages is large, though several methods beyond
the scope of this paper can be employed to keep the calculations
feasible (Salakhutdinov 2008, Lee and Hastie 2012). Alternatively, as
noted below, some common linear and generalized linear methods can also
be used as computationally efficient approximations to the full network
(Lee and Hastie 2012, Loh and Wainwright 2013).

@@

In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When the number of species is
reasonably small, one can compute exact maximum likelihood estimates for
all of the \(\alpha\) and \(\beta\) coefficients given a
presence-absence matrix by optimizing \(p(\vec{y}; \alpha, \beta)\).
Fully-observed Markov networks like the ones considered here have
unimodal likelihood surfaces (Murphy 2012), ensuring that this procedure
will always converge on the global maximum. This maximum represents the
unique combination of \(\alpha\) and \(\beta\) coefficients that would
be expected to produce exactly the observed co-occurrence frequencies on
average (i.e.~maximizing the likelihood ensures that the sufficient
statistics of the model distribution match the sufficient statistics of
the data; Murphy 2012). I used the rosalia package (Harris 2015a) for
the R programming language (R Core Team 2015) to optimize the Markov
network parameters. The package was named after Santa Rosalia, the
patron saint of biodiversity, whose mythos played an important
rhetorical role in the null model debates of the 1970's and 1980's
(Lewin 1983).

@@

I simulated several sets of landscapes using known parameters as a test
bed for different statistical methods. The first set of landscapes
included the three competing species shown in Figure 1. For each of 1000
replicates, I generated a landscape with 100 sites by sampling from a
probability distribution defined by the figure's interaction
coefficients (Appendix 1). Each of the methods described below was then
evaluated on its ability to correctly infer that the two shrub species
competed with one another, despite their frequent co-occurrence.

I then generated landscapes with up to 20 interacting species at 25,
200, or 1600 sites using three increasingly complex models (50
replicates for each combination of size and model). As described in
Appendix {[}{[}X{]}{]}, I randomly drew the ``true'' coefficient values
for each replicate landscape in these simulations so that most species
pairs interacted negligibly, a few pairs interacted very strongly, and
competition was three times more common than facilitation.

The first set of 20-species landscapes, like the landscapes with three
species, were generated directly from a Markov network to ensure that
the model could recover the parameters used to generate the ``observed''
co-occurrence data. Then, I added two environmental factors that varied
from location to location across the simulated landscapes, and simulated
a new set of co-occurrence data so that species' \(\alpha\) coefficients
depended on these environmental variables. The latter set of simulated
landscapes provide an important test of the methods' ability to
distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces
like abiotic environmental filtering. This task was made especially
difficult because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013). Finally, I simulated the
population dynamics of 20 interacting species, as described in Appendix
{[}{[}X{]}{]}

@@

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, I added a
weak regularizer (equivalent to a logistic prior with scale 2) to ensure
that the model estimates were always finite {[}{[}Appendix{]}{]}.

I also evaluated six alternative methods: five from the existing
literature, plus a novel combination of two of these methods. The first
alternative interaction metric was the sample correlation between
species' presence-absence vectors, which summarizes their marginal
association. Next, I used partial correlations, which summarize species'
conditional relationships. This approach is common in molecular biology
(Friedman et al. 2008), but less so in ecology (see Albrecht and Gotelli
2001 @faisal\_inferring\_2010 for two exceptions). In the context of
non-Gaussian data, the partial correlation can be thought of as a
computationally efficient approximation to the full Markov network model
(Loh and Wainwright 2013). Because partial correlations are undefined
for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on James-Stein shrinkage, as implemented in
the corpcor package's \texttt{pcor.shrink} function with the default
settings (Schäfer et al. 2014).

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network (Lee and
Hastie 2012). Following Faisal et al. (2010), I fit regularized logistic
regression models (Gelman et al. 2008) for each species, using the other
species on the landscape as predictors. This produced two interaction
estimates for each species pair; one for the effect of species \(i\) on
species \(j\) and one for the reverse. These two coefficients are not
identifiable from the data, however (Schmidt and Murphy 2012), so I used
their average as an overall measure of the overall relationship between
species \(i\) and species \(j\).

@@

I then evaluated the relationship between each method's estimates and
the ``true'' interaction strengths among all of the species pairs from
the larger simulated landscapes. This determined which of the methods
provide a consistent way to know how strong species interactions
are---regardless of which species were present in a particular data set
or how many observations were taken. To ensure that the different test
statistics (e.g.~correlations versus \(Z\) scores) were on a common
scale, I rescaled them using linear regression through the origin. After
rescaling each method's estimates, I calculated squared errors between
the scaled interaction estimates and ``true'' interaction values across
the simulated data sets. These squared errors determined the proportion
of variance explained for different combinations of model type and
landscape size (compared with a null model that assumed all interaction
strengths to be zero).

@@

Despite some variability across contexts (Figure 3), the four methods
that controlled for indirect effects (the Markov network, the
generalized linear models, and the two partial correlation-based
methods) clearly outperformed those that did not. Of these, the Markov
network consistently performed the best, especially on the smaller
landscapes (which are also the ones that are most representative of
typical analyses in this field; Gotelli and Ulrich 2009).

Of the methods that did not control for indirect effects, Figure 3 shows
that simple correlation coefficients provided a more reliable indicator
of species' true interaction strengths than either the joint species
distribution model (BayesComm) or the null model (Pairs). Interestingly,
the estimates from these three approaches were tightly correlated
{[}{[}Figure 3B{]}{]}. This suggests that the effect sizes from the null
model (and, to a lesser extent, the correlation matrices from joint
species distribution models) only contain noisy versions of the same
information that could be obtained more easily and interpretably by
calculating correlation coefficients between species' presence-absence
vectors.

Bootstrap resampling indicated that the above ranking of the different
methods was robust (Appendix 3). In particular, the 95\% confidence
interval of the bootstrap distribution indicated that the Markov
network's overall \(R^2\) value was between {[}{[}14 and 18{]}{]}
percent higher than the second-most effective method (generalized linear
models) and between {[}{[}2.12 and 2.38{]}{]} times higher than could be
achieved by the null model (Pairs).

@@

The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where environmental heterogeneity and indirect interactions cause
ecologists' typical null modeling approaches to reliably fail. For cases
where fitting a Markov network is computationally infeasible, these
results also indicate that partial covariances and generalized linear
models can both provide useful approximations. The partial correlations'
success on simulated data may not carry over to real data sets, however;
Loh and Wainwright (2013) show that the linear approximations can be
less reliable in cases where the true interaction matrix contains more
structure (e.g.~guilds or trophic levels). Similarly, the approximation
involved in using separate generalized linear models for each species
can occasionally lead to catastrophic overfitting with small-to-moderate
sample sizes (Lee and Hastie 2012). For these reasons, it will usually
be best to fit a Markov network rather than one of the alternative
methods when one's computational resources allow it.

@@

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century (Cipra 1987), and the
models' properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species
interactions would thus allow ecologists to tap into an enormous set of
existing discoveries and techniques for dealing with indirect effects,
stability, and alternative stable states. Numerous other extensions are
possible: for example, the states of the interaction network can be
modeled as a function of the local abiotic environment (Lee and Hastie
2012), which would provide a rigorous and straightforward approach to
the difficult and important task of incorporating whole networks of
biotic interactions into species distribution models (Pollock et al.
2014), leading to a better understanding of the interplay between biotic
and abiotic effects on community structure. There are even methods
(Whittam and Siegel-Causey 1981, Tjelmeland and Besag 1998) that would
allow one species to affect the sign or strength of the relationship
between two other species, tipping the balance between facilitation and
exploitation (Bruno et al. 2003).

@@

\textbf{Figure 1: A.} A small network of three competing species. The
tree (top) tends not to co-occur with either of the two shrub species,
as indicated by the strongly negative coefficient linking them. The two
shrub species also compete with one another, but more weakly (circled
coefficient). \textbf{B.} In spite of the competitive interactions
between the two shrub species, their shared tendency to occur in
locations without trees makes their occurrence vectors positively
correlated (circled). \textbf{C.} Controlling for the tree species'
presence with a conditional (all-else-equal) approach such as a partial
covariance or a Markov network leads to correct identification of the
negative shrub-shrub interaction (circled).

\textbf{Figure 2: A.} A small Markov network with two species, defined
by its \(\alpha\) and \(\beta\) values. The abiotic environment favors
the occurrence of each species (\(\alpha >0\)), particularly species 2
(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
these two species implies that they co-occur less than expected under
independence. \textbf{B.} These coefficients determine the probabilities
of all four possible presence-absence combinations for Species 1 and
Species 2. \(\alpha_1\) is added to the exponent whenever Species 1 is
present (\(y_1 = 1\)), but not when it is absent (\(y_1 = 0\)).
Similarly, the exponent includes \(\alpha_2\) only when species \(2\) is
present (\(y_2 = 1\)), and \(\beta\) only when both are present
(\(y_1y_2 = 1\)). The normalizing constant \(Z\), ensures that the four
probabilities sum to 1. In this case, \(Z\) is about 18.5. \textbf{C.}
The expected frequencies of all possible co-occurrence patterns between
the two species of interest, as calculated in the previous panel.
\textbf{D.} If, as assumed in null model simulations, \(\beta_{12}\)
equaled zero (e.g.~if the species no longer competed for the same
resources), then the reduction in competition would allow each species
to increase its overall occurrence rate.

\textbf{Figure 3: A.} Proportion of variance in interaction coefficients
explained by each method versus number of sampled locations. \textbf{B.}
The \(Z\)-scores produced by the null model (``Pairs'') for each pair of
species can be predicted using the correlation between the
presence-absence vectors of those same species and from the number of
sites on the landscape.

@@

@@

@@

@@

Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
\end{flushleft}
\end{spacing}
